---
title: "R Review (ECON/POLISCI 151)"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basics
Recommended IDE is [RStudio](https://www.rstudio.com/products/rstudio/)

Useful file types:

* R script: Basic file containing only code (and comments)
* Markdown files: Mix of text and code (e.g., this file was made using markdown)
  + Code from other languages besides R
  + Multiple output options: HTML, pdf, etc.
  + Include LaTeX code for math: $\forall x\in \mathbb{R}^n, c\in \mathbb{R}, f(cx)=c^kf(x)$

## Packages
Install packages from CRAN:
```{r}
## Comment out code using '#'
#install.packages("PackageName")
```

To use the package, you must first attach/load it:
```{r}
#require(PackageName)
# or, equivalently,
#library(PackageName)
```

You can use functions from an installed package without attaching it (also a good way to specify package):
```{r}
#PackageName::function()
#DifferentPackage::function()

# if you do not specify the package from which a function is from, R will assume
# it is from the package last loaded. When loading the package, it will notify you
# that some functions are being "masked"
require(dplyr)
# if you call lag(), R will assume you mean dplyr::lag(). If you want to call
# stats::lag(), you must specify
```

Install packages from github:
```{r}
#devtools::install_github("DeveloperName/PackageName")
```

## Loading Data


## Functions
```{r}
# ?function will open the documentation for function in the "help" pane in RStudio
#?mean

# can also press the tab key inside the parenthesis after a function
mean()
```
"Usage" will give you the order of arguments and their default values (if any).

You can specify function the argument and the value it should take by using <tt>argument=value</tt>, e.g.,

```{r}
set.seed(123)
rnorm(n=5, mean = 1, sd = 10)
```

You can also specify arguments out of order:
```{r}
set.seed(123)
rnorm(mean = 1, sd = 10, n=5)
```

If you do not specify the argument, R will assume they are entered in the order in which they appear in "Usage." 
```{r}
set.seed(123)
rnorm(5, 1, 10)
```

If you do not supply a value for an argument with a default value, R will use the default value. If you do not supply a value when there is no default, R will return an error.
```{r}
rnorm(5)
#rnorm()
```

### Defining your own functions
```{r}
foo <- function(x, y) {
  if ( x < y ) {
    return(x)
  } else {
    return(x + y)
  }
}
foo(10, 2)
foo(y=2, x=10)
```


## Objects and data types
Instantiate new objects:
```{r}
a1 <- 1
print(a1)
a1  # can omit "print()"

# can include multiple commands on one line by separating with ";"
a2 = 2; print(a2)
```

### Logicals/booleans:
```{r}
typeof(TRUE)
T  # TRUE

# checking (in)equality
1 == 1
1 != 1
1 <= 1
1 > 1

# can treat as a number
1+TRUE
1+FALSE
```


### Numbers
```{r}
typeof(1)
typeof(1L)
typeof(as.integer(1))
# "is.integer()" does not check for whole numbers
is.integer(1)
```

Vectors:
```{r}
v1 <- c(1,2,3); v1
v2 <- 1:3; v2

# different data types
typeof(v1); typeof(v2)

# sequence, with step size other than 1
seq(from=1, to=10, by=3)
```

Matrices:
```{r}
## white space does not delimit code; use it for readability
# bad:
mat <- rbind(1:3,           4:6,
             
             

             
                     7:9)
mat
# better:
mat <- rbind(1:3,
             4:6,
             7:9)
mat
typeof(mat)

matrix(1:9, nrow=3, byrow = T)
# Matrix operations:
t(mat)  # transpose
mat * 2  # elementwise multiplication
mat * mat  
mat %*% mat  # matrix multiplication
mat + mat
mat + 1
```

### Characters/strings
```{r}
s1 <- "this is a string"; cat(s1)
typeof(s1)

# some special symbols need to be escaped:
cat("s \ ")
cat("s \\ ")
```

### Factors
Discrete variables
```{r}
f1 <- factor(1:3)
sort(f1)  # default order is alpha-numerical

# can specify different order
f2 <- factor(1:3, levels=c(2, 1, 3))
sort(f2)
```

### Dataframes
```{r}
# can accomodate multiple data types
df <- cbind.data.frame(numbers=1:4, words=as.character(1:4))
df

# if we try to do the same with matrices, everything is coerced to char:
cbind(numbers=1:4, words=as.character(1:4))
```
### Lists
```{r}
# can store basically any data together in one object
list(c("a", "b"), 1:3,  # vectors of different sizes
     lm(1~1))  # other classes of objects
```

## Manipulating data
### Subsetting and appending
Subsetting by location:
```{r}
df[1, 1]  # element in row 1, column 1
df[1, ]  # entire first row
df[, 1]  # entire first column
```

By name:
```{r}
df[, "numbers"]
df$numbers
df[["numbers"]]
```

Appending data:
```{r}
# this does not add a column to the object "df"
# instead, it creates a new object with an additional column
cbind(df, new_column1=df$numbers+1)  
df  # the same as before

# to add the column to "df", you have to assign the new object to the variable "df"
df <- cbind(df, new_column=df$numbers+1)  
df

# equivalently:
df$new_column1 <- df$numbers+1
df
```

Useful package:
```{r}
require(dplyr)  # also tidyr
df <- df %>% mutate(new_column2 = (numbers <= 2),
                    new_column3 = ifelse(numbers <= 2,
                      yes="at most 2", no="greater than 2"))
df

# summarize by different groups
df %>% group_by(new_column3) %>% summarize(avg=mean(numbers))

# maintain original structure
df %>% group_by(new_column3) %>% mutate(avg=mean(numbers))
```


### <tt>apply()</tt>
Collection of functions for applying the same function many times. The <tt>apply()</tt> function applies the same function to each row (<tt>MARGIN=1</tt>) or column (<tt>MARGIN=2</tt>) of a matrix/dataframe.
```{r}
## Take the average of each column:
x <- matrix(rnorm(1000), ncol=10)

# for loop
avgs <- c()
for ( i in 1:ncol(x) ) {
  avgs[i] <- mean(x[, i])
}
avgs

# apply()
apply(x, MARGIN=2, FUN=mean)


## Something a bit more complicated
vals <- c()
for ( i in 1:ncol(x) ) {
  vals[i] <- (x[, i] %*% x[, i])/sd(x[, i])
}
vals 

apply(x, MARGIN=2, FUN=function(xi) xi %*% xi / sd(xi) )
```

Similar function for other data structures:
```{r}
# lapply and sapply can both be used on either a list or a vector
lst <- list( x, x%*%t(x) )
vec <- 1:3

# lapply returns a list
lapply(lst, FUN = mean)
lapply(vec, FUN=function(y) y+1)

# sapply returns a vector
sapply(lst, FUN=mean)
sapply(vec, FUN=function(y) y+1)
```

<tt>tapply()</tt> allows you to apply a function to subsets of a vector:
```{r}
set.seed(123)
group <- c("a", "b")[rbinom(10, size=1, prob=.5)+1]

# averages of entries in groups a and b
tapply(rnorm(10), INDEX = group, mean)

```


## Random variables and probability distributions
Generate random numbers:
```{r}
set.seed(123)  # set seed to ensure reproducibility
runif(n=3, min=0, max=1) # uniform(0, 1)
rnorm(n=3, mean=10, sd=1) # normal(0, 1)
```

Quantiles of a distribution:
```{r}
qnorm(.025, mean=0, sd=1)  # 2.5th quantile of N(0,1)
```

CDF:
```{r}
pnorm(0, mean=0, sd=1) # F(0) = P( N(0,1) < 0 )
```

PDF:
```{r}
dnorm(0, 0, 1) # f(0)
```

## Regression
```{r}
set.seed(123)
x1 <- rnorm(n=10)
x2 <- rnorm(n=10)

Y <- x1 + 2*x2 + rnorm(n=10)

# Estimate the model Y = beta1*x1 + beta2*x2 + error
# can pass lm() vectors
out <- lm(Y ~ x1 + x2)
out$residuals  # model has many things you can extract
summary(out)
coef(out)
confint(out)


# data can also pass it a dataframe
df <- cbind.data.frame(Y, Var1=x1, Var2=x2)
head(df)
fm <- formula(Y~Var1+Var2)
lm(fm, data=df)
lm(Y~., data=df)  # regress Y on all other variables in df

# can transform data
Y <- x1^2 + log(abs(x2)) + rnorm(n=10)
lm(Y ~ Var1^2 + log(abs(Var2)), data=df)

# many other functions take the same formula
df$Y <- rbinom(n=10, size=1, prob=.5)  # binary outcome
glm(fm, data=df, family = binomial(link=probit))  # probit 
```

## Plots
Dot plot:
```{r}
plot(x=x1, y=Y, pch=16,
     xlab="x axis", ylab="y axis")
```

Line
```{r}
plot(x=sort(x1), y=Y[order(x1)], type = "l")
```

Histogram and density
```{r}
x <- rnorm(1000, )
hist(x)
plot(density(x)) 
```
